---
title: "MetaCyc reaction-functional profiling from HUMAnN3 outputs level"
author: "Gloria Sarmina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
---
This R Markdown document performs a publication-oriented functional analysis of HUMAnN3-derived MetaCyc reaction CPM profiles.

The workflow:

1. Creates a new project folder named `MetaCyc`.
2. Imports the HUMAnN3 MetaCyc reaction CPM table.
3. Cleans non-informative and stratified entries.
4. Renames samples according to desert origin.
5. Builds a sample-by-reaction matrix.
6. Assigns reactions to ecological-functional categories.
7. Keeps the final 18 functional categories.
8. Performs PERMANOVA on Bray-Curtis dissimilarities.
9. Generates a publication-style heatmap.
10. Generates violin plots with log10-scaled abundance.
11. Exports statistical and ecological interpretation tables.
12. Writes a logbook documenting why each step was done, what data were used, and what outputs were generated.

The input file is expected here:

```text
/Users/alejandrasarmina/Desktop/HUMAnN3_analisis/rxn-cpm-metacyc.tsv
```

The output folder will be:

```text
/Users/alejandrasarmina/Desktop/HUMAnN3_analisis/MetaCyc
```

---

# 1. Project setup and folder architecture

This first step defines the project structure.
A copy of the raw input file is stored inside the project folder to preserve traceability.

```{r setup-project, message=FALSE, warning=FALSE}
# ============================================================
# 1. Clean R session
# ============================================================

rm(list = ls())
cat("\014")
options(scipen = 999)

# ============================================================
# 2. Define input and output paths
# ============================================================

base_dir <- "/Users/alejandrasarmina/Desktop/HUMAnN3_analisis"
project_dir <- file.path(base_dir, "MetaCyc")

input_file <- file.path(base_dir, "rxn-cpm-metacyc.tsv")

# ============================================================
# 3. Define folder architecture
# ============================================================

dir_raw        <- file.path(project_dir, "00_raw_input")
dir_logs       <- file.path(project_dir, "01_logbook")
dir_tables     <- file.path(project_dir, "02_processed_tables")
dir_stats      <- file.path(project_dir, "03_statistics")
dir_figures    <- file.path(project_dir, "04_figures")
dir_supplement <- file.path(project_dir, "05_supplementary_tables")

# Create folders if they do not exist
dir.create(project_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_raw, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_logs, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_tables, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_stats, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_figures, recursive = TRUE, showWarnings = FALSE)
dir.create(dir_supplement, recursive = TRUE, showWarnings = FALSE)

# ============================================================
# 4. Check input file
# ============================================================

if (!file.exists(input_file)) {
  stop("Input file not found: ", input_file)
}

# Copy the raw file into the project folder
file.copy(
  from = input_file,
  to = file.path(dir_raw, basename(input_file)),
  overwrite = TRUE
)

# ============================================================
# 5. Create logbook
# ============================================================

logbook_file <- file.path(dir_logs, "MetaCyc_analysis_logbook.md")

cat(
  paste0(
    "# MetaCyc analysis logbook\n\n",
    "Date: ", Sys.Date(), "\n\n",
    "Input file: `", input_file, "`\n\n",
    "Project folder: `", project_dir, "`\n\n",
    "## Folder architecture\n\n",
    "- `00_raw_input/`: copy of the original HUMAnN3 MetaCyc reaction CPM table.\n",
    "- `01_logbook/`: written record of the analytical workflow.\n",
    "- `02_processed_tables/`: cleaned and transformed data tables.\n",
    "- `03_statistics/`: PERMANOVA and category-level statistical outputs.\n",
    "- `04_figures/`: publication-quality figures in JPG and PDF format.\n",
    "- `05_supplementary_tables/`: supplementary tables for manuscript use.\n\n"
  ),
  file = logbook_file
)

# Function to add entries to the logbook
log_step <- function(step_title, why, input_data, output_data) {
  cat(
    paste0(
      "\n## ", step_title, "\n\n",
      "**Why this step was done:** ", why, "\n\n",
      "**Input data:** ", input_data, "\n\n",
      "**Output generated:** ", output_data, "\n\n"
    ),
    file = logbook_file,
    append = TRUE
  )
}

log_step(
  step_title = "Project folder architecture was created",
  why = "A structured folder system improves reproducibility and separates raw files, processed tables, statistical outputs, figures and logbook records.",
  input_data = "Original HUMAnN3 MetaCyc reaction CPM table.",
  output_data = "MetaCyc project folder with organized subfolders."
)

# Show folder architecture in the HTML output
folder_architecture <- data.frame(
  Folder = c(
    "00_raw_input",
    "01_logbook",
    "02_processed_tables",
    "03_statistics",
    "04_figures",
    "05_supplementary_tables"
  ),
  Purpose = c(
    "Stores a copy of the original HUMAnN3 MetaCyc reaction CPM table.",
    "Stores the written analysis logbook.",
    "Stores cleaned and transformed tables.",
    "Stores statistical outputs.",
    "Stores publication-quality figures.",
    "Stores supplementary tables for manuscript use."
  )
)

folder_architecture
```

---

# 2. Load required libraries

The workflow uses packages for data manipulation, ecological statistics, heatmap construction and publication-quality plotting.

The package `ggh4x` is optional.
If it is available, it allows each violin plot facet to have its own log10 y-axis with several tick values.
If it is not available, the protocol automatically falls back to a shared fixed y-axis.

```{r load-libraries, message=FALSE, warning=FALSE}
# ============================================================
# 1. Define required packages
# ============================================================

cran_repo <- "https://cloud.r-project.org"

required_packages <- c(
  "tidyverse",
  "vegan",
  "pheatmap",
  "ggplot2",
  "scales"
)

# ============================================================
# 2. Install missing required packages
# ============================================================

packages_to_install <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(packages_to_install) > 0) {
  install.packages(packages_to_install, repos = cran_repo)
}

# ============================================================
# 3. Load required packages
# ============================================================

for (pkg in required_packages) {
  library(pkg, character.only = TRUE)
}

# grid is part of base/recommended R packages and is used for pheatmap layout editing
library(grid)

# ============================================================
# 4. Optional package: ggh4x
# ============================================================

if (!"ggh4x" %in% rownames(installed.packages())) {
  message("Optional package ggh4x is not installed. Trying to install it...")
  try(
    install.packages("ggh4x", repos = cran_repo),
    silent = TRUE
  )
}

ggh4x_available <- "ggh4x" %in% rownames(installed.packages())

if (ggh4x_available) {
  library(ggh4x)
  message("ggh4x is available. Violin plots will use independent y-axis scales per facet.")
} else {
  message("ggh4x is not available. Violin plots will use a fixed shared y-axis scale.")
}

log_step(
  step_title = "Required R packages were loaded",
  why = "The workflow requires packages for data manipulation, ecological statistics, heatmaps and publication-quality plotting.",
  input_data = "Installed R packages in the local R environment.",
  output_data = "Libraries loaded into the R session. ggh4x is used if available."
)

sessionInfo()
```

---

# 3. Read HUMAnN3 MetaCyc reaction CPM table

The input table corresponds to MetaCyc reaction-level abundances generated from HUMAnN3.

Because the file name includes `cpm`, the values are interpreted as CPM-normalized MetaCyc reaction abundances.

```{r read-input-table, message=FALSE, warning=FALSE}
# ============================================================
# 1. Import the MetaCyc reaction CPM table
# ============================================================

metacyc_raw <- readr::read_tsv(
  input_file,
  show_col_types = FALSE
)

# Save original name of first column
original_feature_col <- colnames(metacyc_raw)[1]

# Rename first column for consistency
colnames(metacyc_raw)[1] <- "MetaCyc_reaction"

# ============================================================
# 2. Record raw dimensions
# ============================================================

raw_dimensions <- data.frame(
  Metric = c("Number of rows", "Number of columns"),
  Value = c(nrow(metacyc_raw), ncol(metacyc_raw))
)

write_csv(
  raw_dimensions,
  file.path(dir_tables, "01_raw_table_dimensions.csv")
)

log_step(
  step_title = "Raw MetaCyc reaction CPM table was imported",
  why = "The HUMAnN3 reaction-level CPM table is the starting point for all functional profiling analyses.",
  input_data = paste0("File: ", input_file),
  output_data = paste0(
    "Raw table loaded as metacyc_raw. Original feature column was named `",
    original_feature_col,
    "`."
  )
)

raw_dimensions
```

---

# 4. Clean non-informative and stratified entries

HUMAnN3 tables may include entries such as `UNMAPPED` and `UNINTEGRATED`.
These are useful quality-control categories, but they do not represent interpretable MetaCyc reactions.

Stratified rows containing `|` are also removed because this analysis focuses on total community-level functional profiles, not taxon-specific contributions.

```{r clean-metacyc-table, message=FALSE, warning=FALSE}
# ============================================================
# 1. Remove non-informative entries
# ============================================================

n_initial <- nrow(metacyc_raw)

metacyc_clean <- metacyc_raw %>%
  filter(!is.na(MetaCyc_reaction)) %>%
  filter(MetaCyc_reaction != "") %>%
  filter(!str_detect(MetaCyc_reaction, regex("UNMAPPED|UNINTEGRATED", ignore_case = TRUE)))

n_after_unmapped <- nrow(metacyc_clean)

# ============================================================
# 2. Keep only unstratified reactions
# ============================================================

metacyc_unstratified <- metacyc_clean %>%
  filter(!str_detect(MetaCyc_reaction, "\\|"))

n_after_unstratified <- nrow(metacyc_unstratified)

# ============================================================
# 3. Save cleaning summary
# ============================================================

cleaning_summary <- data.frame(
  Step = c(
    "Initial rows",
    "After removing UNMAPPED/UNINTEGRATED/empty rows",
    "After keeping only unstratified reactions"
  ),
  Number_of_rows = c(
    n_initial,
    n_after_unmapped,
    n_after_unstratified
  )
)

write_csv(
  cleaning_summary,
  file.path(dir_tables, "02_cleaning_summary.csv")
)

write_csv(
  metacyc_unstratified,
  file.path(dir_tables, "03_metacyc_unstratified_clean_table.csv")
)

log_step(
  step_title = "Non-informative and stratified MetaCyc entries were removed",
  why = "UNMAPPED and UNINTEGRATED entries do not represent interpretable reactions. Stratified rows were removed to focus on total community-level functional profiles.",
  input_data = "Raw MetaCyc reaction CPM table.",
  output_data = "Clean unstratified MetaCyc reaction table and cleaning summary."
)

cleaning_summary
```

---

# 5. Rename samples according to the experimental design

The HUMAnN3 output columns are renamed to make the ecological design explicit.

CC = Cuatro Ciénegas
VG = Valle de Guadalupe

```{r rename-samples, message=FALSE, warning=FALSE}
# ============================================================
# 1. Extract original sample columns
# ============================================================

sample_columns_original <- colnames(metacyc_unstratified)[-1]

# Remove common HUMAnN3 abundance suffixes
sample_names_clean <- sample_columns_original %>%
  str_remove("_Abundance-RPKs$") %>%
  str_remove("_Abundance-CPM$") %>%
  str_remove("_CPM$") %>%
  str_remove("_RPKs$")

# ============================================================
# 2. Define sample renaming dictionary
# ============================================================

sample_dictionary <- c(
  "A1"  = "A1_CC",
  "A21" = "A2_CC",
  "A22" = "A3_CC",
  "A4"  = "B1_CC",
  "A5"  = "B2_CC",
  "A6"  = "B3_CC",
  "A7"  = "C1_CC",
  "A8"  = "C2_CC",
  "A9"  = "C3_CC",
  "A23" = "D1_VG",
  "A11" = "D2_VG",
  "A12" = "D3_VG",
  "A13" = "E1_VG",
  "A14" = "E2_VG",
  "A15" = "E3_VG",
  "A16" = "F1_VG",
  "A17" = "F2_VG",
  "A24" = "F3_VG"
)

new_sample_names <- sample_dictionary[sample_names_clean]

# Safety check
if (any(is.na(new_sample_names))) {
  missing_names <- sample_names_clean[is.na(new_sample_names)]
  stop(
    "Some sample names were not found in the renaming dictionary: ",
    paste(missing_names, collapse = ", ")
  )
}

# ============================================================
# 3. Rename columns
# ============================================================

metacyc_renamed <- metacyc_unstratified
colnames(metacyc_renamed)[-1] <- new_sample_names

sample_name_table <- data.frame(
  Original_column = sample_columns_original,
  Clean_sample_name = sample_names_clean,
  Renamed_sample = new_sample_names
)

write_csv(
  sample_name_table,
  file.path(dir_tables, "04_sample_renaming_dictionary.csv")
)

write_csv(
  metacyc_renamed,
  file.path(dir_tables, "05_metacyc_clean_renamed_table.csv")
)

log_step(
  step_title = "Samples were renamed according to desert origin and sampling layout",
  why = "Renaming samples makes the ecological design explicit and enables downstream grouping by desert.",
  input_data = "Clean unstratified MetaCyc reaction table with original HUMAnN3 sample names.",
  output_data = "Renamed MetaCyc reaction table and sample renaming dictionary."
)

sample_name_table
```

---

# 6. Build sample-by-reaction matrix and metadata table

Ecological analyses require a matrix with samples as rows and functions as columns.

```{r build-functional-matrix, message=FALSE, warning=FALSE}
# ============================================================
# 1. Build sample-by-reaction matrix
# ============================================================

metacyc_matrix <- metacyc_renamed %>%
  column_to_rownames("MetaCyc_reaction") %>%
  t() %>%
  as.data.frame()

# Ensure all values are numeric
metacyc_matrix <- metacyc_matrix %>%
  mutate(across(everything(), as.numeric))

# Replace missing values with zero
metacyc_matrix[is.na(metacyc_matrix)] <- 0

# Remove reactions with zero abundance across all samples
reaction_sums <- colSums(metacyc_matrix, na.rm = TRUE)
metacyc_matrix <- metacyc_matrix[, reaction_sums > 0, drop = FALSE]

# ============================================================
# 2. Build sample metadata
# ============================================================

metadata_samples <- data.frame(
  Sample = rownames(metacyc_matrix),
  Desert = case_when(
    str_detect(rownames(metacyc_matrix), "_CC$") ~ "CC",
    str_detect(rownames(metacyc_matrix), "_VG$") ~ "VG",
    TRUE ~ "Unknown"
  )
)

metadata_samples$Desert_full <- recode(
  metadata_samples$Desert,
  "CC" = "Cuatro Ciénegas",
  "VG" = "Valle de Guadalupe"
)

rownames(metadata_samples) <- metadata_samples$Sample

# ============================================================
# 3. Save matrix and metadata
# ============================================================

write_csv(
  metacyc_matrix %>% rownames_to_column("Sample"),
  file.path(dir_tables, "06_metacyc_sample_by_reaction_matrix.csv")
)

write_csv(
  metadata_samples,
  file.path(dir_tables, "07_sample_metadata.csv")
)

matrix_summary <- data.frame(
  Metric = c(
    "Number of samples",
    "Number of retained MetaCyc reactions",
    "Number of CC samples",
    "Number of VG samples"
  ),
  Value = c(
    nrow(metacyc_matrix),
    ncol(metacyc_matrix),
    sum(metadata_samples$Desert == "CC"),
    sum(metadata_samples$Desert == "VG")
  )
)

write_csv(
  matrix_summary,
  file.path(dir_tables, "08_matrix_summary.csv")
)

log_step(
  step_title = "Sample-by-reaction matrix and metadata table were generated",
  why = "Ecological analyses require samples as rows and functional features as columns. Metadata are required to compare functional profiles between deserts.",
  input_data = "Clean renamed MetaCyc reaction table.",
  output_data = "Sample-by-reaction matrix, sample metadata table and matrix summary."
)

matrix_summary
```

---

# 7. Assign reactions to ecological-functional categories

MetaCyc reaction identifiers or names can be difficult to interpret directly.
This step uses keyword matching to classify reactions into broad ecological-functional categories.

This classification is heuristic and should be interpreted as an ecological annotation layer, not as an official MetaCyc ontology.

```{r assign-functional-categories, message=FALSE, warning=FALSE}
# ============================================================
# 1. Function to assign categories by keyword matching
# ============================================================

assign_functional_category <- function(reaction_name) {
  
  x <- reaction_name
  
  case_when(
    str_detect(x, regex("amino acid degradation|degradation of arginine|degradation of leucine|degradation of valine|degradation of tyrosine|degradation of tryptophan|degradation of lysine", ignore_case = TRUE)) ~
      "Amino acid degradation",
    
    str_detect(x, regex("arginine|leucine|lysine|histidine|tryptophan|methionine|valine|isoleucine|phenylalanine|tyrosine|cysteine|serine|threonine|alanine|glutamate|glutamine|aspartate|asparagine|proline|amino acid biosynthesis", ignore_case = TRUE)) ~
      "Amino acid biosynthesis",
    
    str_detect(x, regex("biofilm|adhesin|adhesion|pili|fimbriae|fimbria|extracellular polymeric", ignore_case = TRUE)) ~
      "Biofilm and adhesion",
    
    str_detect(x, regex("superoxide|catalase|peroxidase|oxidative|reactive oxygen|heat shock|cold shock|osmotic|stress|environmental response", ignore_case = TRUE)) ~
      "Oxidative stress and environmental response",
    
    str_detect(x, regex("fermentation|lactate|butyrate|ethanol|alcohol|acetate fermentation|mixed acid", ignore_case = TRUE)) ~
      "Fermentation",
    
    str_detect(x, regex("photosynthesis|photosystem|chlorophyll|bacteriochlorophyll|light-harvesting|light harvesting|carotenoid|retinal|phototrophy", ignore_case = TRUE)) ~
      "Photosynthesis and pigments",
    
    str_detect(x, regex("glycolysis|EMP|Entner|Doudoroff|gluconeogenesis|TCA|tricarboxylic|pyruvate|acetyl|citrate|succinate|fumarate|malate|glyoxylate|central carbon", ignore_case = TRUE)) ~
      "Glycolysis and central metabolism",
    
    str_detect(x, regex("benzoate|phenol|catechol|aromatic|toluene|xylene|salicylate|protocatechuate", ignore_case = TRUE)) ~
      "Aromatic compound metabolism",
    
    str_detect(x, regex("sulfate|sulfite|sulfur|sulphur|sulfide|sulphide|thiosulfate|cysteine sulfur|sulfonate", ignore_case = TRUE)) ~
      "Sulfur metabolism",
    
    str_detect(x, regex("phosphate|phosphonate|phosphatase|polyphosphate|phosphorus|phosphoribosyl", ignore_case = TRUE)) ~
      "Phosphorus metabolism",
    
    str_detect(x, regex("methane|methanogenesis|methylamine|methanol|formate|formaldehyde|one-carbon|one carbon|C1 metabolism", ignore_case = TRUE)) ~
      "Methane and C1 metabolism",
    
    str_detect(x, regex("nitrate|nitrite|ammonium|ammonia|nitrogenase|\\bnif|denitrification|nitrification|urea|urease|nitrogen fixation|nitric oxide|nitrous oxide", ignore_case = TRUE)) ~
      "Nitrogen metabolism",
    
    str_detect(x, regex("purine|pyrimidine|adenine|guanine|cytosine|thymine|uracil|nucleotide|nucleoside", ignore_case = TRUE)) ~
      "Nucleotides",
    
    str_detect(x, regex("peptidoglycan|capsule|LPS|lipopolysaccharide|teichoic|cell wall|polysaccharide|exopolysaccharide|EPS|glycan", ignore_case = TRUE)) ~
      "Cell wall and polysaccharides",
    
    str_detect(x, regex("UV|ultraviolet|photolyase|DNA repair|mycosporine|melanin|sunscreen|photoprotection", ignore_case = TRUE)) ~
      "UV protection",
    
    str_detect(x, regex("cytochrome|oxidase|reductase|ferredoxin|quinone|menaquinone|ubiquinone|electron transport|respiration|NADH dehydrogenase|ATP synthase", ignore_case = TRUE)) ~
      "Respiration and electron transport",
    
    str_detect(x, regex("biotin|folate|cobalamin|thiamine|riboflavin|cofactor|vitamin|NAD|FAD|heme|pantothenate|pyridoxal|menaquinone|ubiquinone", ignore_case = TRUE)) ~
      "Vitamins and cofactors",
    
    TRUE ~ "Other or unclassified functions"
  )
}

# ============================================================
# 2. Apply annotation function to all reactions
# ============================================================

annotation_reactions <- data.frame(
  Functional_category = sapply(colnames(metacyc_matrix), assign_functional_category)
)

rownames(annotation_reactions) <- colnames(metacyc_matrix)

# ============================================================
# 3. Define final 18 categories
# ============================================================

selected_categories_final <- c(
  "Phosphorus metabolism",
  "Other or unclassified functions",
  "Fermentation",
  "Respiration and electron transport",
  "Aromatic compound metabolism",
  "Amino acid biosynthesis",
  "UV protection",
  "Nucleotides",
  "Vitamins and cofactors",
  "Nitrogen metabolism",
  "Methane and C1 metabolism",
  "Sulfur metabolism",
  "Cell wall and polysaccharides",
  "Biofilm and adhesion",
  "Amino acid degradation",
  "Glycolysis and central metabolism",
  "Oxidative stress and environmental response",
  "Photosynthesis and pigments"
)

# Legend order used in the heatmap
heatmap_category_order <- c(
  "Amino acid biosynthesis",
  "Amino acid degradation",
  "Biofilm and adhesion",
  "Oxidative stress and environmental response",
  "Fermentation",
  "Photosynthesis and pigments",
  "Glycolysis and central metabolism",
  "Aromatic compound metabolism",
  "Sulfur metabolism",
  "Phosphorus metabolism",
  "Methane and C1 metabolism",
  "Nitrogen metabolism",
  "Nucleotides",
  "Other or unclassified functions",
  "Cell wall and polysaccharides",
  "UV protection",
  "Respiration and electron transport",
  "Vitamins and cofactors"
)

# Facet order used in the violin plot
violin_category_order <- c(
  "Sulfur metabolism",
  "Oxidative stress and environmental response",
  "Amino acid degradation",
  "Phosphorus metabolism",
  "Other or unclassified functions",
  "UV protection",
  "Amino acid biosynthesis",
  "Nucleotides",
  "Biofilm and adhesion",
  "Respiration and electron transport",
  "Cell wall and polysaccharides",
  "Fermentation",
  "Vitamins and cofactors",
  "Nitrogen metabolism",
  "Aromatic compound metabolism",
  "Photosynthesis and pigments",
  "Methane and C1 metabolism",
  "Glycolysis and central metabolism"
)

# ============================================================
# 4. Save annotation summary
# ============================================================

category_summary_all <- annotation_reactions %>%
  rownames_to_column("MetaCyc_reaction") %>%
  count(Functional_category, name = "Number_of_reactions") %>%
  arrange(desc(Number_of_reactions))

write_csv(
  annotation_reactions %>% rownames_to_column("MetaCyc_reaction"),
  file.path(dir_tables, "09_reaction_functional_category_annotation.csv")
)

write_csv(
  category_summary_all,
  file.path(dir_tables, "10_all_functional_category_summary.csv")
)

log_step(
  step_title = "MetaCyc reactions were assigned to ecological-functional categories",
  why = "Reaction-level outputs were grouped into interpretable ecological-functional categories to support figure generation and statistical comparisons.",
  input_data = "MetaCyc reaction identifiers or names.",
  output_data = "Reaction annotation table and category summary."
)

category_summary_all
```

---

# 8. Select the final 18 ecological-functional categories

This step restricts the analysis to the final set of categories used in the publication-style heatmap and violin plot.

```{r select-final-categories, message=FALSE, warning=FALSE}
# ============================================================
# 1. Select reactions belonging to the final 18 categories
# ============================================================

selected_reactions_publication <- rownames(annotation_reactions)[
  annotation_reactions$Functional_category %in% selected_categories_final
]

if (length(selected_reactions_publication) == 0) {
  stop("No reactions matched the final selected ecological-functional categories.")
}

metacyc_matrix_selected <- metacyc_matrix[, selected_reactions_publication, drop = FALSE]

annotation_reactions_selected <- annotation_reactions[
  selected_reactions_publication,
  ,
  drop = FALSE
] %>%
  mutate(
    Functional_category = factor(
      Functional_category,
      levels = heatmap_category_order
    )
  )

# ============================================================
# 2. Save selected category summary
# ============================================================

selected_category_summary <- annotation_reactions_selected %>%
  rownames_to_column("MetaCyc_reaction") %>%
  count(Functional_category, name = "Number_of_reactions") %>%
  arrange(Functional_category)

missing_final_categories <- setdiff(
  selected_categories_final,
  as.character(unique(annotation_reactions_selected$Functional_category))
)

missing_category_table <- data.frame(
  Missing_category = missing_final_categories
)

write_csv(
  selected_category_summary,
  file.path(dir_tables, "11_final_selected_category_summary.csv")
)

write_csv(
  annotation_reactions_selected %>% rownames_to_column("MetaCyc_reaction"),
  file.path(dir_tables, "12_final_selected_reactions.csv")
)

write_csv(
  missing_category_table,
  file.path(dir_tables, "13_missing_final_categories_if_any.csv")
)

log_step(
  step_title = "Final 18 ecological-functional categories were selected",
  why = "The selected categories reproduce the target publication figures and ensure consistency between the heatmap, violin plots and statistical outputs.",
  input_data = "Reaction category annotation and sample-by-reaction CPM matrix.",
  output_data = "Selected reaction matrix, selected reaction annotation, category summary and missing-category report."
)

if (length(missing_final_categories) > 0) {
  warning(
    "The following final categories had no matched reactions and will not appear in plots: ",
    paste(missing_final_categories, collapse = ", ")
  )
}

selected_category_summary
```

---

# 9. PERMANOVA using Bray-Curtis dissimilarity

PERMANOVA tests whether the overall MetaCyc reaction profile differs between deserts.

Bray-Curtis dissimilarity is calculated from relative abundances of the selected MetaCyc reactions.

```{r permanova-functional-profiles, message=FALSE, warning=FALSE}
# ============================================================
# 1. Convert CPM matrix to relative abundance
# ============================================================

metacyc_relative_selected <- metacyc_matrix_selected / rowSums(metacyc_matrix_selected)
metacyc_relative_selected[is.na(metacyc_relative_selected)] <- 0

# ============================================================
# 2. Bray-Curtis dissimilarity
# ============================================================

bray_metacyc <- vegan::vegdist(
  metacyc_relative_selected,
  method = "bray"
)

# ============================================================
# 3. PERMANOVA
# ============================================================

permanova_metacyc <- vegan::adonis2(
  bray_metacyc ~ Desert,
  data = metadata_samples,
  permutations = 999
)

permanova_table <- as.data.frame(permanova_metacyc) %>%
  rownames_to_column("Term")

write_csv(
  permanova_table,
  file.path(dir_stats, "01_PERMANOVA_MetaCyc_18_categories_BrayCurtis.csv")
)

# Label used in the heatmap
r2_perm <- permanova_metacyc$R2[1]
p_perm <- permanova_metacyc$`Pr(>F)`[1]

permanova_label <- paste0(
  "PERMANOVA: R² = ",
  round(r2_perm, 3),
  ", p = ",
  signif(p_perm, 3)
)

log_step(
  step_title = "PERMANOVA was performed on selected MetaCyc functional profiles",
  why = "PERMANOVA tests whether overall selected MetaCyc reaction profiles differ between deserts.",
  input_data = "Relative abundance matrix restricted to reactions from the final 18 ecological-functional categories and sample metadata.",
  output_data = "Bray-Curtis distance object and PERMANOVA table."
)

permanova_table
```

---

# 10. Publication heatmap

This heatmap reproduces the publication-style figure using the final 18 ecological-functional categories.

The heatmap is built at the reaction level.
Rows are samples and columns are MetaCyc reactions.
Each reaction is scaled across samples using a z-score.

```{r heatmap-publication, message=FALSE, warning=FALSE, fig.width=14, fig.height=10}
# ============================================================
# 1. Scale reaction abundances by function
# ============================================================

metacyc_scaled_selected <- scale(metacyc_matrix_selected)

# ============================================================
# 2. Prepare sample annotations
# ============================================================

annotation_samples <- metadata_samples %>%
  select(Desert)

# ============================================================
# 3. Define colors for categories and desert
# ============================================================

category_palette_selected <- c(
  "#C9A400",
  "#12B8C9",
  "#27B5FF",
  "#59C52C",
  "#D98CE8",
  "#F7A6F1",
  "#D77ACF",
  "#F27AA8",
  "#4AA5E8",
  "#F2AA00",
  "#E97BDA",
  "#B9C300",
  "#8BCF00",
  "#27C7F2",
  "#15C95A",
  "#9C95F0",
  "#F5A25C",
  "#20D3C2"
)

category_colors_selected <- setNames(
  category_palette_selected,
  heatmap_category_order
)

annotation_colors_selected <- list(
  Desert = c(
    "CC" = "#F4A09C",
    "VG" = "#19C9A3"
  ),
  Functional_category = category_colors_selected
)

# ============================================================
# 4. Generate heatmap
# ============================================================

heatmap_selected_categories <- pheatmap::pheatmap(
  metacyc_scaled_selected,
  annotation_row = annotation_samples,
  annotation_col = annotation_reactions_selected,
  annotation_colors = annotation_colors_selected,
  show_colnames = FALSE,
  show_rownames = FALSE,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  fontsize = 9,
  border_color = NA,
  color = colorRampPalette(c("navy", "white", "red2"))(100),
  main = "MetaCyc reactions grouped by ecological category\nFunction-wise scaled abundance (z-score)",
  silent = TRUE
)

# Display heatmap in the knitted document
grid::grid.newpage()
grid::grid.draw(heatmap_selected_categories$gtable)

grid::grid.text(
  permanova_label,
  x = 0.5,
  y = 0.02,
  gp = grid::gpar(fontsize = 11, fontface = "bold")
)

# ============================================================
# 5. Save JPG
# ============================================================

jpeg(
  filename = file.path(dir_figures, "Figure_01_MetaCyc_18_categories_heatmap_zscore_PERMANOVA.jpg"),
  width = 14,
  height = 10,
  units = "in",
  res = 300
)

grid::grid.newpage()
grid::grid.draw(heatmap_selected_categories$gtable)

grid::grid.text(
  permanova_label,
  x = 0.5,
  y = 0.02,
  gp = grid::gpar(fontsize = 13, fontface = "bold")
)

dev.off()

# ============================================================
# 6. Save PDF
# ============================================================

pdf(
  file = file.path(dir_figures, "Figure_01_MetaCyc_18_categories_heatmap_zscore_PERMANOVA.pdf"),
  width = 14,
  height = 10
)

grid::grid.newpage()
grid::grid.draw(heatmap_selected_categories$gtable)

grid::grid.text(
  permanova_label,
  x = 0.5,
  y = 0.02,
  gp = grid::gpar(fontsize = 13, fontface = "bold")
)

dev.off()

log_step(
  step_title = "Publication heatmap was generated",
  why = "The heatmap reproduces the target figure by grouping MetaCyc reactions into the final ecological-functional categories and scaling abundance by reaction.",
  input_data = "Selected MetaCyc reaction CPM matrix, sample metadata, reaction category annotation and PERMANOVA results.",
  output_data = "Publication heatmap in JPG and PDF format."
)
```

---

# 11. Build long-format table for category-level abundance

To reproduce the violin plot, MetaCyc reaction abundances are summed within each sample and ecological-functional category.

This produces one abundance value per sample per category.

```{r build-category-abundance-table, message=FALSE, warning=FALSE}
# ============================================================
# 1. Prepare selected reaction annotation
# ============================================================

annotation_reactions_selected_clean <- annotation_reactions_selected %>%
  rownames_to_column("MetaCyc_reaction") %>%
  mutate(
    Functional_category = as.character(Functional_category)
  )

# ============================================================
# 2. Convert selected matrix to long format
# ============================================================

metacyc_long_selected <- metacyc_matrix_selected %>%
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  pivot_longer(
    cols = -Sample,
    names_to = "MetaCyc_reaction",
    values_to = "CPM_abundance"
  ) %>%
  left_join(
    metadata_samples %>% select(Sample, Desert, Desert_full),
    by = "Sample"
  ) %>%
  left_join(
    annotation_reactions_selected_clean,
    by = "MetaCyc_reaction"
  ) %>%
  filter(Functional_category %in% selected_categories_final)

# ============================================================
# 3. Sum CPM abundance per sample and category
# ============================================================

metacyc_category_sample_selected <- metacyc_long_selected %>%
  group_by(Sample, Desert, Desert_full, Functional_category) %>%
  summarise(
    Category_abundance_CPM = sum(CPM_abundance, na.rm = TRUE),
    Number_of_reactions = n(),
    .groups = "drop"
  ) %>%
  mutate(
    # A pseudocount of 1 is added only for log10 visualization
    Category_abundance_CPM_plot = Category_abundance_CPM + 1,
    Functional_category = factor(
      Functional_category,
      levels = violin_category_order
    )
  )

# ============================================================
# 4. Summary by desert
# ============================================================

category_summary_by_desert_selected <- metacyc_category_sample_selected %>%
  group_by(Functional_category, Desert) %>%
  summarise(
    Mean_abundance_CPM = mean(Category_abundance_CPM, na.rm = TRUE),
    Median_abundance_CPM = median(Category_abundance_CPM, na.rm = TRUE),
    SD = sd(Category_abundance_CPM, na.rm = TRUE),
    N = n(),
    SE = SD / sqrt(N),
    .groups = "drop"
  )

# ============================================================
# 5. Save outputs
# ============================================================

write_csv(
  metacyc_long_selected,
  file.path(dir_tables, "14_metacyc_long_selected_18_categories_CPM.csv")
)

write_csv(
  metacyc_category_sample_selected,
  file.path(dir_tables, "15_metacyc_selected_18_categories_by_sample_CPM.csv")
)

write_csv(
  category_summary_by_desert_selected,
  file.path(dir_tables, "16_selected_18_category_summary_by_desert_CPM.csv")
)

log_step(
  step_title = "Category-level abundance table was generated",
  why = "To reproduce the target violin plot, MetaCyc reaction abundances were summed within each sample and ecological-functional category.",
  input_data = "Selected MetaCyc reaction CPM matrix, sample metadata and selected reaction annotations.",
  output_data = "Long-format table, category-by-sample abundance table and desert-level category summary."
)

head(metacyc_category_sample_selected)
```

---

# 12. Statistical comparison between deserts

Each ecological-functional category is compared between deserts using a Wilcoxon rank-sum test.

P-values are corrected using the Benjamini-Hochberg false discovery rate method.

Cliff's delta is reported as a non-parametric effect size.

```{r category-statistics, message=FALSE, warning=FALSE}
# ============================================================
# 1. Function to calculate Cliff's delta
# ============================================================

cliffs_delta_simple <- function(x, y) {
  x <- x[!is.na(x)]
  y <- y[!is.na(y)]
  
  if (length(x) == 0 | length(y) == 0) {
    return(NA_real_)
  }
  
  comparisons <- outer(x, y, "-")
  
  delta <- (sum(comparisons > 0) - sum(comparisons < 0)) /
    (length(x) * length(y))
  
  return(delta)
}

# ============================================================
# 2. Wilcoxon test, FDR correction and effect size
# ============================================================

category_statistics_selected <- metacyc_category_sample_selected %>%
  mutate(Functional_category = as.character(Functional_category)) %>%
  group_by(Functional_category) %>%
  summarise(
    mean_CC = mean(Category_abundance_CPM[Desert == "CC"], na.rm = TRUE),
    mean_VG = mean(Category_abundance_CPM[Desert == "VG"], na.rm = TRUE),
    median_CC = median(Category_abundance_CPM[Desert == "CC"], na.rm = TRUE),
    median_VG = median(Category_abundance_CPM[Desert == "VG"], na.rm = TRUE),
    delta_mean_VG_minus_CC = mean_VG - mean_CC,
    
    p_value = wilcox.test(
      Category_abundance_CPM[Desert == "CC"],
      Category_abundance_CPM[Desert == "VG"],
      exact = FALSE
    )$p.value,
    
    cliffs_delta = cliffs_delta_simple(
      Category_abundance_CPM[Desert == "VG"],
      Category_abundance_CPM[Desert == "CC"]
    ),
    
    .groups = "drop"
  ) %>%
  mutate(
    q_value = p.adjust(p_value, method = "BH"),
    
    significance = case_when(
      q_value < 0.001 ~ "***",
      q_value < 0.01  ~ "**",
      q_value < 0.05  ~ "*",
      TRUE ~ ""
    ),
    
    effect_size_magnitude = case_when(
      abs(cliffs_delta) < 0.147 ~ "negligible",
      abs(cliffs_delta) < 0.33  ~ "small",
      abs(cliffs_delta) < 0.474 ~ "medium",
      abs(cliffs_delta) >= 0.474 ~ "large",
      TRUE ~ NA_character_
    ),
    
    direction = case_when(
      cliffs_delta > 0 ~ "VG > CC",
      cliffs_delta < 0 ~ "CC > VG",
      cliffs_delta == 0 ~ "No clear direction",
      TRUE ~ NA_character_
    ),
    
    Functional_category = factor(
      Functional_category,
      levels = violin_category_order
    )
  ) %>%
  arrange(Functional_category)

write_csv(
  category_statistics_selected,
  file.path(dir_stats, "02_selected_18_categories_Wilcoxon_FDR_CliffsDelta.csv")
)

log_step(
  step_title = "Statistical tests were performed for the final 18 ecological-functional categories",
  why = "Wilcoxon rank-sum tests were used to compare summed category-level abundance between deserts. Benjamini-Hochberg correction controls the false discovery rate.",
  input_data = "Summed CPM abundance per sample and ecological-functional category.",
  output_data = "Wilcoxon p-values, FDR-corrected q-values, Cliff's delta, effect-size magnitude and direction of enrichment."
)

category_statistics_selected
```

---

# 13. Violin plot with log10 abundance axis

This figure reproduces the target violin plot.

Each panel represents one ecological-functional category.
The y-axis is displayed in log10 scale.
Asterisks indicate categories that remain significant after FDR correction.

If `ggh4x` is available, each facet receives its own y-axis breaks, with at least four tick values whenever possible.
If `ggh4x` is not available, the plot uses a shared fixed y-axis.

```{r violin-plot-log10, message=FALSE, warning=FALSE, fig.width=16, fig.height=10}
# ============================================================
# 1. Add significance asterisks to facet labels
# ============================================================

category_panel_labels_selected <- category_statistics_selected %>%
  mutate(
    Functional_category = as.character(Functional_category),
    Category_panel = if_else(
      significance != "",
      paste0(Functional_category, " *"),
      Functional_category
    )
  ) %>%
  select(Functional_category, Category_panel)

metacyc_category_panel_selected <- metacyc_category_sample_selected %>%
  mutate(Functional_category = as.character(Functional_category)) %>%
  left_join(category_panel_labels_selected, by = "Functional_category")

panel_order_selected <- category_statistics_selected %>%
  mutate(
    Functional_category = as.character(Functional_category),
    Category_panel = if_else(
      significance != "",
      paste0(Functional_category, " *"),
      Functional_category
    )
  ) %>%
  arrange(match(Functional_category, violin_category_order)) %>%
  pull(Category_panel)

metacyc_category_panel_selected <- metacyc_category_panel_selected %>%
  mutate(
    Category_panel = factor(Category_panel, levels = panel_order_selected)
  )

# ============================================================
# 2. Function to create at least four y-axis breaks per facet
# ============================================================

candidate_breaks <- c(
  1, 3, 10, 30, 100, 300,
  1000, 3000, 10000, 30000, 100000, 300000
)

get_log_breaks <- function(x) {
  x <- x[is.finite(x) & !is.na(x) & x > 0]
  
  if (length(x) == 0) {
    return(c(1, 3, 10, 30))
  }
  
  min_x <- min(x)
  max_x <- max(x)
  
  idx_low_candidates <- which(candidate_breaks <= min_x)
  idx_high_candidates <- which(candidate_breaks >= max_x)
  
  idx_low <- ifelse(
    length(idx_low_candidates) > 0,
    max(idx_low_candidates),
    1
  )
  
  idx_high <- ifelse(
    length(idx_high_candidates) > 0,
    min(idx_high_candidates),
    length(candidate_breaks)
  )
  
  while ((idx_high - idx_low + 1) < 4) {
    if (idx_low > 1) {
      idx_low <- idx_low - 1
    }
    if ((idx_high - idx_low + 1) >= 4) break
    if (idx_high < length(candidate_breaks)) {
      idx_high <- idx_high + 1
    }
    if (idx_low == 1 && idx_high == length(candidate_breaks)) break
  }
  
  candidate_breaks[idx_low:idx_high]
}

# ============================================================
# 3. Create individual y-axis scales for facets when ggh4x exists
# ============================================================

facet_breaks_table <- metacyc_category_panel_selected %>%
  group_by(Category_panel) %>%
  summarise(
    breaks_list = list(get_log_breaks(Category_abundance_CPM_plot)),
    .groups = "drop"
  ) %>%
  arrange(match(Category_panel, panel_order_selected))

facet_scales_y <- lapply(facet_breaks_table$breaks_list, function(bks) {
  scale_y_log10(
    breaks = bks,
    limits = range(bks),
    labels = scales::comma_format(accuracy = 1)(bks),
    minor_breaks = NULL
  )
})

# ============================================================
# 4. Base violin plot
# ============================================================

base_violin_plot <- ggplot(
  metacyc_category_panel_selected,
  aes(
    x = Desert,
    y = Category_abundance_CPM_plot,
    fill = Desert
  )
) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    color = NA,
    alpha = 0.6
  ) +
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA,
    alpha = 0.9,
    color = "#333333"
  ) +
  scale_fill_manual(
    values = c(
      "CC" = "#F4A09C",
      "VG" = "#68D5D2"
    ),
    labels = c(
      "CC" = "CC",
      "VG" = "VG"
    )
  ) +
  labs(
    title = NULL,
    x = "Desert",
    y = "Abundance (log10)",
    caption = paste(
      "Wilcoxon rank-sum tests were used to compare category-level summed CPM abundance between deserts.",
      "P-values were adjusted using the Benjamini-Hochberg false discovery rate correction.",
      "Asterisks indicate q < 0.05 after FDR correction.",
      "A pseudocount of 1 was added only for visualization on the log10 scale.",
      sep = "\n"
    )
  ) +
  theme_minimal(base_size = 16) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 30, hjust = 1, size = 12),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.position = "none",
    plot.caption = element_text(hjust = 0, size = 10),
    panel.grid.minor = element_blank()
  )

# ============================================================
# 5. Final violin plot
# ============================================================

if (ggh4x_available) {
  
  figure_violin_selected_categories <- base_violin_plot +
    facet_wrap(
      ~ Category_panel,
      scales = "free_y",
      ncol = 4
    ) +
    ggh4x::facetted_pos_scales(
      y = facet_scales_y
    )
  
} else {
  
  figure_violin_selected_categories <- base_violin_plot +
    facet_wrap(
      ~ Category_panel,
      scales = "fixed",
      ncol = 4
    ) +
    scale_y_log10(
      breaks = c(
        1, 3, 10, 30, 100, 300,
        1000, 3000, 10000, 30000, 100000, 300000
      ),
      labels = c(
        "1", "3", "10", "30", "100", "300",
        "1,000", "3,000", "10,000", "30,000", "100,000", "300,000"
      )
    )
}

figure_violin_selected_categories

# ============================================================
# 6. Save violin plot
# ============================================================

ggsave(
  filename = file.path(dir_figures, "Figure_02_MetaCyc_18_categories_violin_log10.jpg"),
  plot = figure_violin_selected_categories,
  width = 16,
  height = 10,
  units = "in",
  dpi = 300
)

ggsave(
  filename = file.path(dir_figures, "Figure_02_MetaCyc_18_categories_violin_log10.pdf"),
  plot = figure_violin_selected_categories,
  width = 16,
  height = 10,
  units = "in"
)

log_step(
  step_title = "Violin plot was generated for the final 18 ecological-functional categories",
  why = "The figure reproduces the target publication layout by showing category-level abundance distributions on a log10 scale and marking statistically significant categories.",
  input_data = "Summed CPM abundance per sample and ecological-functional category, plus Wilcoxon/FDR results.",
  output_data = "Publication-style violin plot in JPG and PDF format."
)
```

---

# 14. Ecological interpretation table

This table links statistical results to ecological interpretation for the same 18 categories used in the heatmap and violin plot.

```{r ecological-interpretation-table, message=FALSE, warning=FALSE}
# ============================================================
# 1. Calculate dominant desert per category
# ============================================================

category_wide_selected <- metacyc_category_sample_selected %>%
  mutate(Functional_category = as.character(Functional_category)) %>%
  group_by(Functional_category, Desert) %>%
  summarise(
    Mean_abundance_CPM = mean(Category_abundance_CPM, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = Desert,
    values_from = Mean_abundance_CPM
  ) %>%
  mutate(
    Dominant_desert = case_when(
      CC > VG ~ "Cuatro Ciénegas",
      VG > CC ~ "Valle de Guadalupe",
      TRUE ~ "Similar"
    )
  )

# ============================================================
# 2. Merge statistics with ecological interpretation
# ============================================================

ecological_interpretation_selected <- category_statistics_selected %>%
  mutate(Functional_category = as.character(Functional_category)) %>%
  left_join(category_wide_selected, by = "Functional_category") %>%
  mutate(
    Statistical_support = if_else(
      q_value < 0.05,
      "Supported after FDR correction",
      "Not significant after FDR correction"
    ),
    
    Ecological_interpretation = case_when(
      q_value < 0.05 & Dominant_desert == "Cuatro Ciénegas" ~
        "Higher category-level abundance in Cuatro Ciénegas.",
      
      q_value < 0.05 & Dominant_desert == "Valle de Guadalupe" ~
        "Higher category-level abundance in Valle de Guadalupe.",
      
      q_value >= 0.05 ~
        "No statistically supported difference after FDR correction.",
      
      TRUE ~ "Pattern requires further inspection."
    ),
    
    Functional_category = factor(
      Functional_category,
      levels = violin_category_order
    )
  ) %>%
  arrange(Functional_category)

write_csv(
  ecological_interpretation_selected,
  file.path(dir_supplement, "Supplementary_Table_MetaCyc_18_categories_ecological_interpretation.csv")
)

log_step(
  step_title = "Ecological interpretation table was generated",
  why = "Statistical results were translated into ecological interpretation using the same 18 functional categories shown in the publication figures.",
  input_data = "Category-level statistics and desert-level mean abundances.",
  output_data = "Supplementary ecological interpretation table."
)

ecological_interpretation_selected
```

---

# 15. Final output summary

This final section records all files generated by the workflow.

```{r final-output-summary, message=FALSE, warning=FALSE}
# ============================================================
# 1. List generated files
# ============================================================

generated_files <- list.files(
  project_dir,
  recursive = TRUE,
  full.names = TRUE
)

generated_files_table <- data.frame(
  File = generated_files,
  Folder = dirname(generated_files),
  File_name = basename(generated_files)
)

write_csv(
  generated_files_table,
  file.path(project_dir, "MetaCyc_generated_files_summary.csv")
)

log_step(
  step_title = "Analysis completed",
  why = "A final output summary helps identify all files generated by the workflow and supports reproducibility.",
  input_data = "All files generated inside the MetaCyc project directory.",
  output_data = "MetaCyc_generated_files_summary.csv and completed analysis logbook."
)

generated_files_table
```

---

# 16. Main outputs

After running the workflow, the main results will be saved in:

```text
/Users/alejandrasarmina/Desktop/HUMAnN3_analisis/MetaCyc
```

The main figures will be saved in:

```text
/Users/alejandrasarmina/Desktop/HUMAnN3_analisis/MetaCyc/04_figures
```

Main figure files:

```text
Figure_01_MetaCyc_18_categories_heatmap_zscore_PERMANOVA.jpg
Figure_01_MetaCyc_18_categories_heatmap_zscore_PERMANOVA.pdf
Figure_02_MetaCyc_18_categories_violin_log10.jpg
Figure_02_MetaCyc_18_categories_violin_log10.pdf
```

The logbook will be saved in:

```text
/Users/alejandrasarmina/Desktop/HUMAnN3_analisis/MetaCyc/01_logbook/MetaCyc_analysis_logbook.md
```






